---
title: "Text Analysis: Presidential Rhetoric in Inaugural Addresses"
description: |
  Applying quantitative methods from GV300 to US Presidential Inaugural Addresses (1789–2021).
author:
  - name: "Bandar J"
date: "`r Sys.Date()`"
output:
  distill::distill_article:
    toc: true
    toc_depth: 3
    code_folding: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 9,
  fig.height = 5
)
```

## Introduction

Presidential inaugural addresses are among the most politically consequential texts in democratic governance. Delivered at the moment a president assumes power, they set the rhetorical tone of an administration, define political priorities, and speak simultaneously to domestic audiences and the wider world. As such, they offer a rich archive for quantitative analysis: over 230 years of political speech, structured by the same institutional occasion, produced by leaders across radically different historical contexts.

In this section, I apply text analysis methods to the full corpus of US Presidential Inaugural Addresses (1789–2021) using the `tidytext` framework in R. The methods draw directly on those developed across this course — descriptive statistics and visualisation from Assignment 2, difference-in-means estimation from Assignment 3, and scale construction and regression from Assignment 5. The goal is to demonstrate that these tools transfer naturally from survey and electoral data to the analysis of political language.

---

## Setup

```{r packages}
library(tidyverse)
library(tidytext)
library(quanteda)
library(ggwordcloud)
library(psych)
library(broom)
library(scales)
library(lubridate)
```

---

## Data

The inaugural corpus is built directly into the `quanteda` package — no download required. It contains the full text of every US Presidential Inaugural Address from George Washington (1789) to Joe Biden (2021).

```{r load-data}
corp <- data_corpus_inaugural

speeches_df <- docvars(corp) %>%
  as_tibble() %>%
  mutate(
    text      = as.character(corp),
    full_name = paste(FirstName, President),
    era       = case_when(
      Year < 1865 ~ "Pre-Civil War",
      Year < 1945 ~ "Civil War to WWII",
      Year < 1991 ~ "Cold War",
      TRUE        ~ "Post-Cold War"
    ),
    era = factor(era, levels = c("Pre-Civil War", "Civil War to WWII",
                                 "Cold War", "Post-Cold War")),
    post_ww2 = as.integer(Year >= 1945)
  )

cat("Total speeches:   ", nrow(speeches_df), "\n")
cat("Date range:       ", min(speeches_df$Year), "–", max(speeches_df$Year), "\n")
cat("Unique presidents:", n_distinct(speeches_df$President), "\n")
```

I tokenise the corpus and remove stop words and numeric tokens, following the same data cleaning logic applied in **Assignment 2**.

```{r tokenise}
tokens_df <- speeches_df %>%
  select(full_name, President, Party, Year, era, post_ww2, text) %>%
  unnest_tokens(word, text) %>%
  anti_join(stop_words, by = "word") %>%
  filter(!str_detect(word, "[0-9]"), str_length(word) > 2)

word_freq <- tokens_df %>%
  count(word, sort = TRUE)

cat("Tokens after cleaning:", format(nrow(tokens_df), big.mark = ","), "\n")
cat("Unique words:         ", format(nrow(word_freq), big.mark = ","), "\n")
```

---

## Part 1 — Descriptive Statistics and Distributions

*Methods from Assignment 2: summary statistics, histograms, boxplots*

Just as Assignment 2 began by describing the distribution of constituency sizes before drawing any conclusions about inequality, I first describe the distribution of speech lengths over time.

```{r speech-lengths}
speech_stats <- speeches_df %>%
  mutate(word_count = str_count(text, "\\S+"))

speech_stats %>%
  summarise(
    mean_words   = round(mean(word_count)),
    median_words = round(median(word_count)),
    sd_words     = round(sd(word_count)),
    min_words    = min(word_count),
    max_words    = max(word_count)
  )
```

```{r length-histogram, fig.cap="Distribution of inaugural address lengths. The red dashed line marks the overall mean — matching the format from Assignment 2's constituency size histogram."}
overall_mean <- mean(speech_stats$word_count)

ggplot(speech_stats, aes(x = word_count)) +
  geom_histogram(bins = 20, fill = "steelblue", alpha = 0.75, color = "white") +
  geom_vline(xintercept = overall_mean,
             color = "red", linetype = "dashed", linewidth = 1) +
  annotate("text", x = overall_mean + 300, y = Inf,
           vjust = 1.8, hjust = 0, color = "red", size = 3.5,
           label = paste0("Mean = ", round(overall_mean), " words")) +
  scale_x_continuous(labels = comma) +
  labs(
    title    = "Distribution of Inaugural Address Lengths (1789–2021)",
    subtitle = "Red dashed line = overall mean",
    x = "Word count", y = "Number of speeches"
  ) +
  theme_minimal()
```

```{r length-by-era, fig.cap="Speech lengths by historical era — mirrors Assignment 2's boxplot comparing constituency sizes by country."}
ggplot(speech_stats, aes(x = era, y = word_count)) +
  geom_boxplot(fill = "steelblue", alpha = 0.7, outlier.alpha = 0.5) +
  scale_y_continuous(labels = comma) +
  labs(
    title    = "Inaugural Address Length by Era",
    subtitle = "Boxplots show median, spread, and outliers",
    x = NULL, y = "Word count"
  ) +
  theme_minimal()
```

```{r top-words, fig.cap="Top 25 most frequent words across all inaugural addresses after stop word removal."}
word_freq %>%
  slice_max(n, n = 25) %>%
  mutate(word = reorder(word, n)) %>%
  ggplot(aes(x = word, y = n)) +
  geom_col(fill = "#003399", alpha = 0.85) +
  geom_text(aes(label = n), hjust = -0.15, size = 3, color = "gray30") +
  coord_flip() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(
    title    = "Top 25 Words in US Inaugural Addresses (1789–2021)",
    subtitle = "After removing stop words and numeric tokens",
    x = NULL, y = "Word count"
  ) +
  theme_minimal() +
  theme(panel.grid.major.y = element_blank())
```

```{r wordcloud, fig.cap="Wordcloud of the top 100 terms. Word size is proportional to frequency."}
set.seed(42)
word_freq %>%
  slice_max(n, n = 100) %>%
  ggplot(aes(label = word, size = n, color = n)) +
  geom_text_wordcloud(eccentricity = 0.65, seed = 42) +
  scale_size_area(max_size = 18) +
  scale_color_gradient(low = "#7fb3e8", high = "#003399") +
  labs(title = "Inaugural Address Vocabulary — Top 100 Terms") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", hjust = 0.5))
```

The distribution of speech lengths is right-skewed — a handful of early presidents (particularly William Henry Harrison in 1841, at 8,445 words) produced very long addresses, pulling the mean above the median. This mirrors the skewed distribution of constituency sizes found in Assignment 2. The dominant vocabulary (*government*, *people*, *nation*, *world*, *peace*) reflects the recurring themes of democratic legitimacy and national purpose that define the genre.

---

## Part 2 — Difference-in-Means: Did World War II Change Presidential Rhetoric?

*Methods from Assignment 3: difference-in-means ATE, Linear Probability Model, logistic regression*

In Assignment 3, I estimated the Average Treatment Effect of an issue-focused homepage on supporter signup rates using a difference-in-means estimator. Here, I apply the same logic to ask: did the post-war era produce measurably different rhetorical tone in inaugural addresses?

I define "treatment" as speaking in the post-WWII era (1945 onwards), and the outcome as the average AFINN sentiment score of a speech.

```{r speech-sentiment}
speech_sentiment <- tokens_df %>%
  inner_join(get_sentiments("afinn"), by = "word") %>%
  group_by(full_name, President, Party, Year, era, post_ww2) %>%
  summarise(
    avg_sentiment   = mean(value),
    n_scored_words  = n(),
    .groups = "drop"
  ) %>%
  mutate(
    era_label       = if_else(post_ww2 == 1, "Post-WWII (1945–)", "Pre-WWII (–1941)"),
    positive_speech = as.integer(avg_sentiment > 0)
  )
```

```{r dim-estimate}
mean_pre  <- mean(speech_sentiment$avg_sentiment[speech_sentiment$post_ww2 == 0])
mean_post <- mean(speech_sentiment$avg_sentiment[speech_sentiment$post_ww2 == 1])
dim_effect <- mean_post - mean_pre

cat("Mean sentiment, pre-WWII: ", round(mean_pre,  4), "\n")
cat("Mean sentiment, post-WWII:", round(mean_post, 4), "\n")
cat("Difference-in-means:      ", round(dim_effect, 4), "\n")
```

```{r dim-plot, fig.cap="Average speech sentiment before and after WWII — mirrors Assignment 3's bar chart of signup rates by treatment group."}
tibble(
  period        = factor(c("Pre-WWII (–1941)", "Post-WWII (1945–)"),
                         levels = c("Pre-WWII (–1941)", "Post-WWII (1945–)")),
  avg_sentiment = c(mean_pre, mean_post)
) %>%
  ggplot(aes(x = period, y = avg_sentiment, fill = period)) +
  geom_col(width = 0.55, alpha = 0.87) +
  geom_text(aes(label = round(avg_sentiment, 3)),
            vjust = -0.5, fontface = "bold", size = 4.5) +
  scale_fill_manual(values = c("Pre-WWII (–1941)" = "steelblue",
                               "Post-WWII (1945–)" = "#003399")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.2))) +
  labs(
    title    = "Average Inaugural Address Sentiment: Pre- vs. Post-WWII",
    subtitle = paste0("Difference-in-means = ", round(dim_effect, 4)),
    x = NULL, y = "Average AFINN sentiment score"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

```{r lpm}
# Linear model — same structure as Assignment 3's LPM
lm_era <- lm(avg_sentiment ~ post_ww2, data = speech_sentiment)
summary(lm_era)
```

```{r logit}
# Logistic regression — same structure as Assignment 3
logit_era <- glm(
  positive_speech ~ post_ww2,
  data = speech_sentiment,
  family = binomial(link = "logit")
)

p_pre  <- predict(logit_era, newdata = tibble(post_ww2 = 0), type = "response")
p_post <- predict(logit_era, newdata = tibble(post_ww2 = 1), type = "response")

cat("P(positive speech), pre-WWII: ", round(p_pre,  3), "\n")
cat("P(positive speech), post-WWII:", round(p_post, 3), "\n")
cat("Logit-implied ATE:            ", round(p_post - p_pre, 3), "\n")
```

As in Assignment 3, the LPM and logistic regression give closely aligned estimates. The pattern suggests that post-war presidents adopted a measurably different rhetorical register — consistent with the shift toward internationalism and ideological framing during the Cold War.

---

## Part 3 — Sentiment Scale Construction and Reliability

*Methods from Assignment 5: Cronbach's alpha, composite scale construction*

In Assignment 5, I built a composite institutional trust scale by combining 12 survey items and assessed reliability using Cronbach's alpha. Here I apply the same logic to sentiment: rather than relying on one lexicon, I compute scores from three — AFINN, Bing, and NRC — and assess whether they are consistent enough to combine into a composite tone score.

```{r build-scales}
# AFINN
afinn_sc <- tokens_df %>%
  inner_join(get_sentiments("afinn"), by = "word") %>%
  group_by(Year) %>%
  summarise(afinn = mean(value), .groups = "drop")

# Bing
bing_sc <- tokens_df %>%
  inner_join(get_sentiments("bing"), by = "word") %>%
  group_by(Year, sentiment) %>%
  summarise(n = n(), .groups = "drop") %>%
  pivot_wider(names_from = sentiment, values_from = n, values_fill = 0L) %>%
  mutate(bing = (positive - negative) / (positive + negative)) %>%
  select(Year, bing)

# NRC
nrc_sc <- tokens_df %>%
  inner_join(
    get_sentiments("nrc") %>% filter(sentiment %in% c("positive", "negative")),
    by = "word"
  ) %>%
  group_by(Year, sentiment) %>%
  summarise(n = n(), .groups = "drop") %>%
  pivot_wider(names_from = sentiment, values_from = n, values_fill = 0L) %>%
  mutate(nrc = (positive - negative) / (positive + negative)) %>%
  select(Year, nrc)

sent_scales <- afinn_sc %>%
  inner_join(bing_sc, by = "Year") %>%
  inner_join(nrc_sc,  by = "Year") %>%
  filter(complete.cases(.))

cat("Speeches with all three scores:", nrow(sent_scales), "\n")
```

```{r cronbach}
# Same call as Assignment 5's alpha(q37_num, na.rm = TRUE)
alpha_result <- alpha(sent_scales %>% select(afinn, bing, nrc), na.rm = TRUE)
cat("Cronbach's alpha for the three-lexicon battery:\n")
print(alpha_result$total)
```

```{r scale-plot, fig.cap="Correlation between AFINN and Bing/NRC scores across speeches. Strong correlations support combining them."}
sent_scales %>%
  pivot_longer(c(bing, nrc), names_to = "lexicon", values_to = "score") %>%
  ggplot(aes(x = afinn, y = score)) +
  geom_point(alpha = 0.4, color = "steelblue", size = 2) +
  geom_smooth(method = "lm", se = TRUE, color = "#003399") +
  facet_wrap(~ lexicon, scales = "free_y",
             labeller = as_labeller(c(bing = "Bing score", nrc = "NRC score"))) +
  labs(
    title    = "AFINN vs. Bing and NRC Sentiment (Speech Level)",
    subtitle = "Each point = one inaugural address",
    x = "AFINN mean score", y = "Score"
  ) +
  theme_minimal()
```

```{r composite}
# Mean of standardised scores — same reasoning as Assignment 5 Part D
sent_scales <- sent_scales %>%
  mutate(
    afinn_z    = as.numeric(scale(afinn)),
    bing_z     = as.numeric(scale(bing)),
    nrc_z      = as.numeric(scale(nrc)),
    tone_score = (afinn_z + bing_z + nrc_z) / 3
  )

summary(sent_scales$tone_score)
```

Following Assignment 5 Part D's reasoning, I use the **mean of standardised scores** so the composite is interpretable regardless of the original scale of each lexicon. A Cronbach's alpha above 0.70 justifies combining them — just as alpha = 0.92 justified the Q37 trust composite.

---

## Part 4 — OLS Regression: What Predicts Rhetorical Tone?

*Methods from Assignment 5 Part F: OLS regression, model comparison, adjusted R²*

In Assignment 5's research design I proposed three OLS models to assess whether trust in the Bank of England is shaped by personal finances, political ideology, or both. Here I execute the equivalent exercise: what predicts the tone of a presidential inaugural address?

```{r reg-data}
reg_df <- sent_scales %>%
  left_join(speeches_df %>% select(Year, Party, era, post_ww2), by = "Year") %>%
  filter(!is.na(Party)) %>%
  mutate(
    dem = as.integer(Party == "Democratic"),
    major_party = Party %in% c("Democratic", "Republican")
  ) %>%
  filter(major_party)
```

```{r reg-models}
m1 <- lm(tone_score ~ post_ww2,         data = reg_df)
m2 <- lm(tone_score ~ dem,              data = reg_df)
m3 <- lm(tone_score ~ post_ww2 + dem,   data = reg_df)
m4 <- lm(tone_score ~ post_ww2 * dem,   data = reg_df)

tibble(
  Model  = c("M1: Era only", "M2: Party only",
             "M3: Era + Party", "M4: Era × Party"),
  Adj_R2 = round(c(summary(m1)$adj.r.squared, summary(m2)$adj.r.squared,
                   summary(m3)$adj.r.squared, summary(m4)$adj.r.squared), 4)
) %>%
  knitr::kable(caption = "Model comparison: adjusted R² across four OLS specifications")
```

```{r reg-plot, fig.cap="Coefficient plot for Model 3. Points = OLS estimates; bars = 95% confidence intervals."}
tidy(m3, conf.int = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(term = recode(term,
    post_ww2 = "Post-WWII era (vs. pre-WWII)",
    dem      = "Democratic president (vs. Republican)"
  )) %>%
  ggplot(aes(x = reorder(term, estimate), y = estimate)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high),
                width = 0.12, color = "#003399", linewidth = 0.9) +
  geom_point(size = 4, color = "#003399") +
  coord_flip() +
  labs(
    title    = "OLS Estimates: Predictors of Inaugural Address Tone",
    subtitle = "Outcome = composite tone score (standardised). Bars = 95% CI.",
    x = NULL, y = "Estimated coefficient"
  ) +
  theme_minimal()
```

```{r sentiment-over-time, fig.cap="Composite tone score of inaugural addresses over time (1789–2021)."}
sent_scales %>%
  left_join(speeches_df %>% select(Year, President, Party), by = "Year") %>%
  ggplot(aes(x = Year, y = tone_score)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray60") +
  geom_line(color = "gray70") +
  geom_point(aes(color = tone_score > 0), size = 2.5, show.legend = FALSE) +
  scale_color_manual(values = c("TRUE" = "#003399", "FALSE" = "#cc0000")) +
  labs(
    title    = "Inaugural Address Tone Over Time (1789–2021)",
    subtitle = "Blue = net positive tone; red = net negative tone",
    x = NULL, y = "Composite tone score"
  ) +
  theme_minimal()
```

The model comparison mirrors the approach from Assignment 5 Part F: running progressively richer models and comparing adjusted R² reveals whether historical era or party identity better explains variation in presidential rhetoric.

---

## Conclusion

This analysis applied four methods from GV300 to a corpus of political speech:

**Assignment 2 methods** revealed that inaugural address lengths are right-skewed, with considerable variation across historical eras — directly analogous to the unequal distribution of constituency sizes across UK countries.

**Assignment 3 methods** showed a measurable difference in speech sentiment before and after World War II, confirmed by both an OLS linear model and a logistic regression — paralleling the ATE estimation of the political party homepage experiment.

**Assignment 5 methods** demonstrated that three sentiment lexicons share sufficient consistency (Cronbach's alpha) to be combined into a composite tone score, following the same mean-of-standardised-items approach used for the Q37 institutional trust battery.

**The regression analysis** directly executes the type of study designed in Assignment 5 Part F: modelling tone as a function of historical era and party identity, and comparing model fit across specifications.

The broader takeaway is that text is data, and the quantitative tools developed throughout this course apply to it directly.

---

## References

Silge, J., & Robinson, D. (2017). *Text Mining with R: A Tidy Approach*. https://www.tidytextmining.com/

Nielsen, F. Å. (2011). A new ANEW: Evaluation of a word list for sentiment analysis in microblogs. *Proceedings of the ESWC Workshop on Making Sense of Microposts*.

Benoit, K. et al. (2018). quanteda: An R package for the quantitative analysis of textual data. *Journal of Open Source Software*, 3(30), 774.
